A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-09-04 jacobian_calc_CMAME.tex
Added eqn. nums., added ts cal meth exprns

The scope of the application component of the study was limited to the Linear Peridynamic Solid
material model implemented in the computational peridynamics code, \textit{Peridigm}. Identifying a
specific application provided a practical framework for implementing the new method and solving
engineering problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen cause it combined several helpful characteristics: The reliance on
Newton's method and tangent-stiffness matrices for solving quasi-static implicit problems, prior
inclusion of finite-difference, central-difference and automatic-differentiation based methods
needed for comparison to a new method that would be developed, and being largely based in the Ansi
standard version of the popular C++ programming language.

@@

How is a tangent-stiffness matrix computed? The mathematical formula for a tangent stiffness matrix
can be expressed in indicial notation as:\\\\ $\frac{\partial F_i}{\partial X_j}\mid_{X_0,...,X_n}$
\\\\ $F_i$ is the i'th component of the vector valued function, $X_j$ is the j'th component of the
vector variable, and a particular value of the vector variable is chosen $X_0,...,X_n$. One then
evaluates the expression for each combination $i, j$ corresponding to an element at $row, column$ in
the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be
estimated using the CTSE, AD and finite-difference techniques for functions $F:R^1 \rightarrow R^1$,
since taking partial derivatives entails holding all but a single independent variable of the
function constant, and each component of a vector valued function can be evaluated independently of
the other components. \\

For reference, expressions for calculating tangent stiffness matrices with CTSE, AD,
central-difference and forward-difference will be shown and explained.

@@

Once the selection of what to measure, why to measured it, and if it could be measured was
determined, the task was to find a way to take measurements that satisfied our goals. The goals of
collecting accuracy and speed measurements were achieved by developing and implementing metrics
within the simulation program used in the study. The metric used to measure the accuracy of a
tangent-stiffness matrix was the Frobenius norm of the element-wise difference between the
tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix
produced by the AD based method, given that both methods were set upon the same problem. The lower
the value of this metric, the closer the other method's Jacobian was to the AD Jacobian, and
therefore the more accurate the method was. The expression for the accuracy metric:\\ $D =
\sqrt{\sum_{i=0}^{n-1}\sum_{j=0}^{n-1} (J_{AD}(i,j) - J_{M}(i,j))^2}$\\ Where $D$ is distance, n is
the number of degrees of freedom, $J_{AD}$ is the Jacobian matrix produced by the AD based method, M
stands for other \emph{M}ethod, and the root of the summed squared element-wise differences between
the matrices is taken.  The metric used to compare the speed of the methods was the time in seconds
taken to resume execution in the calling method after instructing the implementation of the method
being evaluated to run and complete itself. Another basis of comparison used in the study, but not
quite a metric, called computational efficiency is based on the ratio of calculation time to number
of Jacobian matrix elements.

>>>>

The scope of the application component of the study was limited to the Linear Peridynamic Solid
material model implemented in the computational peridynamics code, \emph{Peridigm}. Identifying a
specific application provided a practical framework for implementing the new method and solving
engineering problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen cause it combined several helpful characteristics: The reliance on
Newton's method and tangent-stiffness matrices for solving quasi-static implicit problems, prior
inclusion of finite-difference, central-difference and automatic-differentiation based methods
needed for comparison to a new method that would be developed, and being largely based in the Ansi
standard version of the popular C++ programming language.

@@

How is a tangent-stiffness matrix computed? The mathematical formula for a tangent stiffness matrix
can be expressed in indicial notation as:

\begin{equation} \frac{\partial F_i}{\partial X_j}\mid_{X_0,...,X_n} \end{equation}

Where $F_i$ is the i'th component of the vector valued function, $X_j$ is the j'th component of the
vector variable, and a particular value of the vector variable is chosen $X_0,...,X_n$. One then
evaluates the expression for each combination $i, j$ corresponding to an element at $row, column$ in
the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be
estimated using the CTSE, AD and finite-difference techniques for functions $F:R^1 \rightarrow R^1$,
since taking partial derivatives entails holding all but a single independent variable of the
function constant, and each component of a vector valued function can be evaluated independently of
the other components. \\

\subsubsection{Calculating Tangent-stiffness Matrices with Each Method}
For reference, expressions for calculating tangent stiffness matrices with CTSE, AD,
central-difference and forward-difference will be shown and explained.

When the using the forward-difference technique the formula for calculating a tangent-stiffness
matrix is:

\begin{equation} K_{ij} = \frac{\partial F_i^{int}(x + \delta e_j) - F_i^{int}(x)}{\delta}
\end{equation}

Where $K_{ij}$ is the index notation representation of an element of the tangent-stiffness matrix at
row $i$, column $j$. The expression says that the partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the quotient of the partial difference of two terms
and a denomenator. The first of the 'two terms' is $F_i$ evaluated at $x$ where
the $j$'th component of $x$ has been perturbed by an very small amount in the direction of the
positive $x_j$ axis. The second of the 'two terms' is $F_I$ evaluated at an unperturbed $x$. The
denominator is the magnitude of perturbation called "probe distance" {ADDAGIO MANUAL REF}

When using the central-difference technique the formula for calculating a tangent-stiffness matrix
is:

\begin{equation} K_{ij} = \frac{\partial F_i^{int}(x + \delta e_j) - F_i^{int}(x - \delta
e_j)}{2\delta} \end{equation}

The distiction between the central difference-method and the forward-difference method
tangent-stiffness calculation formula is that the partial difference is now between two perturbed
function evaluations, rather than between an unperturbed and perturbed function evaluation as in
forward-difference. Additionally, the minuend involves a perturbation of $x$ in the direction opposite to
the direction of the $j$'th component of $x$.

When using the comple-step technique, which is based on CTSE, the formula for calculating a
tangent-stiffness matrix is:

\begin{equation} K_{ij} = \frac{imag(F_i^{int}(x + \delta e_j^{imag}))}{\delta} \end{equation}

Where our model $F$ is treated as function of complex variables and $x$ is treated as a vector of
complex values. The expressions says that the real valued partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the imaginary component of $F_i$ evaluated at $x$
where $x$ is perturbed by a very small amount in the direciton of the imaginary axis corresponding
to the $j$'th component of $x$ all divided by the magnitude of the perturbation. There is no partial
notation because a partial difference is not taken.

When using automatic-differentiation to calculate a tangent-stiffness matrix, one only needs to
follow the definition of the tangent-stiffness matrix and issue the correct commands to the AD
system. The formula is just as expected:

\begin{equation} K_{ij} = \frac{\partial F_i^{int}}{\partial x_j} \end{equation}

@@

Once the selection of what to measure, why to measured it, and if it could be measured was
determined, the task was to find a way to take measurements that satisfied our goals. The goals of
collecting accuracy and speed measurements were achieved by developing and implementing metrics
within the simulation program used in the study. The metric used to measure the accuracy of a
tangent-stiffness matrix was the Frobenius norm of the element-wise difference between the
tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix
produced by the AD based method, given that both methods were set upon the same problem. The lower
the value of this metric, the closer the other method's Jacobian was to the AD Jacobian, and
therefore the more accurate the method was. The expression for the accuracy metric:

\begin{equation} D = \sqrt{\sum_{i=0}^{n-1}\sum_{j=0}^{n-1} (J_{AD}(i,j) - J_{M}(i,j))^2}
\end{equation}

Where $D$ is distance, n is the number of degrees of freedom, $J_{AD}$ is the Jacobian matrix
produced by the AD based method, M stands for other \emph{M}ethod, and the root of the summed
squared element-wise differences between the matrices is taken.  The metric used to compare the
speed of the methods was the time in seconds taken to resume execution in the calling method after
instructing the implementation of the method being evaluated to run and complete itself. Another
basis of comparison used in the study, but not quite a metric, called computational efficiency is
based on the ratio of calculation time to number of Jacobian matrix elements.
